---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Generalized hypergeometric function with tunable high precision

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/hypergeo2)](https://CRAN.R-project.org/package=hypergeo2)
[![R-CMD-check](https://github.com/zhuxr11/hypergeo2/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/zhuxr11/hypergeo2/actions/workflows/R-CMD-check.yaml)
[![Download stats](https://cranlogs.r-pkg.org/badges/grand-total/hypergeo2)](https://CRAN.R-project.org/package=hypergeo2)
<!-- badges: end -->

**Package**: [*hypergeo2*](https://github.com/zhuxr11/hypergeo2) `r pkgload::pkg_version()`<br />
**Author**: Xiurui Zhu<br />
**Modified**: `r file.info("README.Rmd")$mtime`<br />
**Compiled**: `r Sys.time()`

The goal of `hypergeo2` is to implement generalized hypergeometric function with tunable high precision. Two floating-point datatypes -- `mpfr_float` and `gmp_float` -- are channeled into this package. The computation is limited to real numbers, since its underlying workhorse [`boost::math::detail::hypergeometric_pFq_checked_series()`](https://www.boost.org/doc/libs/master/boost/math/special_functions/detail/hypergeometric_pFq_checked_series.hpp) from [`BH` package](https://cran.r-project.org/package=BH) implements comparisons (`>` or `<`) between numbers, which are not defined for complex datatypes.

## Installation

### System requirements

If you are building from source (e.g. not installing binaries on Windows), you need to prepare for two system requirements: [GMP](https://gmplib.org/) and [MPFR](https://www.mpfr.org/), to facilitate high-precision floating point types. You may follow the installation instructions from their websites, or try the following commands for quick (default) installation.

* For Windows with rtools40 (or newer) installed, please run:
```bash
pacman -Syu make pkg-config libtool gmp mpfr
```

* For macOS, please run:
```bash
brew install gmp mpfr
```

* For linux, please build and install platform-specific libraries through their `configure` file, or install them using `conda`.

If you are running R from an isolated environment (e.g. `conda`), you need to first activate the environment, and then build the system requirements and the package in the same environment to avoid version conflicts such as: undefined reference to `memcpy@GLIBC#.#.#`.

### The package

If the requirements are installed into their default paths (e.g. without using the `--prefix` option), you are OK to go ahead installing the package in R, as `pkg-config` will take care finding them.

You can install the released version of `hypergeo2` from [CRAN](https://cran.r-project.org/) with:

``` r
install.packages("hypergeo2")
```

Alternatively, you can install the developmental version of `hypergeo2` from [github](https://github.com/) with:

``` r
remotes::install_github("zhuxr11/hypergeo2")
```

However, if the requirements are not installed into their default paths, you may first need to provide configuration arguments/variables to installation paths. Optionally, you may also designate the [maximal series iteration](https://www.boost.org/doc/libs/master/libs/math/doc/html/math_toolkit/pol_ref/iteration_pol.html), default at 10000. To sum up, you may configure the installation in one of the following ways (replacing `<my_gmp_install>`, `<my_mpfr_install>` and `<my_max_iter>` with your paths) before installing the package with `R CMD INSTALL` in bash:

* Set `--configure-args` in `R CMD INSTALL`:
```bash
MY_GMP_INSTALL='<my_gmp_install>'
MY_MPFR_INSTALL='<my_mpfr_install>'
MY_MAX_ITER=<my_max_iter>
R CMD INSTALL hypergeo2 \
  --configure-args="\
    --with-gmp-include=${MY_GMP_INSTALL}/include \
    --with-mpfr-include=${MY_MPFR_INSTALL}/include \
    --with-gmp-lib=${MY_GMP_INSTALL}/lib \
    --with-mpfr-lib=${MY_MPFR_INSTALL}/lib \
    --with-max-iter=${MY_MAX_ITER}\
  "
```

* Set `--configure-vars` in `R CMD INSTALL`:
```bash
MY_GMP_INSTALL='<my_gmp_install>'
MY_MPFR_INSTALL='<my_mpfr_install>'
MY_MAX_ITER=<my_max_iter>
R CMD INSTALL hypergeo2 \
  --configure-vars="\
    GMP_INCLUDE=${MY_GMP_INSTALL}/include \
    MPFR_INCLUDE=${MY_MPFR_INSTALL}/include \
    GMP_LIB=${MY_GMP_INSTALL}/lib \
    MPFR_LIB=${MY_MPFR_INSTALL}/lib \
    MAX_ITER=${MY_MAX_ITER}\
  "
```

## Examples of generalized hypergeometric function with tunable high precision

Sometimes, computing generalized hypergeometric function in double precision is not sufficient, even though we only need 6-8 accurate digits in the results. Here, two floating-point datatypes are provided: `mpfr_float` ("mpfr") and `gmp_float` ("gmp"). By comparison, the "mpfr" backend is safer, since it defines \code{Inf} while the "gmp" backend throws overflow exception. Therefore, `mpfr_float` is used as the default backend.

```{r library, eval=FALSE}
library(hypergeo2)
```

```{r, include=FALSE}
pkgload::load_all()
```

For example, let us compute a generalized hypergeometric function in [Matlab Online](https://matlab.mathworks.com/) and use its value as reference:

```
>> fprintf("%.32g", hypergeom([-28.2 11.8 15.8], [12.8 17.8], 1))
2.7120446907792120783054486094854e-09
```

```{r, include=FALSE}
ref_value <- 2.7120446907792120783054486094854e-09
print_err <- function(x, ref) format(x / ref - 1, digits = 3L)
```

First, we compute the same function with `hypergeo::genhypergeo()` function.

```{r hypergeo}
hypergeo_U <- c(-28.2, 11.8, 15.8)
hypergeo_L <- c(12.8, 17.8)
hypergeo_z <- 1
(hypergeo_res <- hypergeo::genhypergeo(U = hypergeo_U, L = hypergeo_L, z = hypergeo_z))
```

As can be seen, the result is heavily deviated by `r print_err(hypergeo_res, ref_value)` in terms of relative error.

Then, we compute the same function with `hypergeo2::genhypergeo()` function, at default (double) precision.

```{r hypergeo2-default}
(hypergeo2_res_double <- genhypergeo(U = hypergeo_U, L = hypergeo_L, z = hypergeo_z))
```

As can be seen, the result is still deviated by `r print_err(hypergeo2_res_double, ref_value)` in terms of relative error, although much better.

Next, we compute the same function with `hypergeo2::genhypergeo()` function, with a precision of 20 digits from `mpfr` and `gmp` backends, respectively.

```{r hypergeo2-prec20}
(hypergeo2_res_prec20_mpfr <- genhypergeo(U = hypergeo_U, L = hypergeo_L, z = hypergeo_z,
                                          prec = 20L, backend = "mpfr"))
(hypergeo2_res_prec20_gmp <- genhypergeo(U = hypergeo_U, L = hypergeo_L, z = hypergeo_z,
                                         prec = 20L, backend = "gmp"))
```

As can be seen, the result from `gmp` backend is more precise than that from `mpfr`, with the deviation of `mpfr` at `r print_err(hypergeo2_res_prec20_mpfr, ref_value)` and that of `gmp` at `r print_err(hypergeo2_res_prec20_gmp, ref_value)`. This is because that `gmp` usually use higher precision than we set (see [this post](https://www.mpfr.org/faq.html#:~:text=What%20are%20the%20differences%20between%20MPF%20from%20GMP,minimum%20value%20%28MPF%20generally%20uses%20a%20higher%20precision%29)).

Finally, to validate this hypothesis, we further increase the precision to 25 digits.


```{r hypergeo2-prec25}
(hypergeo2_res_prec25_mpfr <- genhypergeo(U = hypergeo_U, L = hypergeo_L, z = hypergeo_z,
                                          prec = 25L, backend = "mpfr"))
(hypergeo2_res_prec25_gmp <- genhypergeo(U = hypergeo_U, L = hypergeo_L, z = hypergeo_z,
                                         prec = 25L, backend = "gmp"))
```

As can be seen, now both results are very close to the reference, with deviations from `mpfr` and `gmp` backends at `r print_err(hypergeo2_res_prec25_mpfr, ref_value)` and `r print_err(hypergeo2_res_prec25_gmp, ref_value)`, respectively.

## Precision vs time of computation

The time of computation at different precision is profiled in this section.

```{r toc, fig.width=7, fig.height=5, fig.dpi=192}
bench_scheme <- expand.grid(prec = c(list(NULL), as.list(seq(16L, 30L))),
                            backend = c("mpfr", "gmp"),
                            stringsAsFactors = FALSE)
bench_fun <- function(prec, backend, quote = FALSE) {
  res <- bquote(genhypergeo(U = hypergeo_U, L = hypergeo_L, z = hypergeo_z,
                            prec = .(prec), backend = .(backend)) / ref_value - 1)
  if (quote == FALSE) {
    eval(res)
  } else {
    res
  }
}
bench_scheme[["err"]] <- mapply(
  FUN = bench_fun,
  prec = bench_scheme[["prec"]],
  backend = bench_scheme[["backend"]],
  USE.NAMES = FALSE,
  SIMPLIFY = TRUE
)
bench_res <- summary(microbenchmark::microbenchmark(
  list = mapply(
    FUN = bench_fun,
    prec = bench_scheme[["prec"]],
    backend = bench_scheme[["backend"]],
    MoreArgs = list(quote = TRUE),
    USE.NAMES = FALSE,
    SIMPLIFY = TRUE
  ),
  times = 100L
))
bench_res <- cbind(bench_scheme, bench_res[colnames(bench_res) %in% "expr" == FALSE])
bench_res[["prec"]] <- as.character(bench_res[["prec"]])
ggplot2::ggplot(
  bench_res,
  ggplot2::aes(x = as.integer(factor(prec, levels = unique(prec))))
) +
  ggplot2::geom_point(ggplot2::aes(y = mean / 1000), color = "blue") +
  ggplot2::geom_line(ggplot2::aes(y = mean / 1000), color = "blue") +
  ggplot2::geom_point(ggplot2::aes(y = (log(abs(err)) + 20) / 16), color = "red") +
  ggplot2::geom_line(ggplot2::aes(y = (log(abs(err)) + 20) / 16), color = "red") +
  ggplot2::scale_x_continuous(breaks = seq_along(unique(bench_res[["prec"]])),
                              minor_breaks = NULL,
                              labels = unique(bench_res[["prec"]])) +
  ggplot2::scale_y_continuous(
    sec.axis = ggplot2::sec_axis(~. * 16 - 20,
                                 breaks = seq(-20, 0, by = 4),
                                 name = "Logarithmic absolute relative error")
  ) +
  ggplot2::facet_wrap(ggplot2::vars(backend), ncol = 1L) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    axis.title.y.left = ggplot2::element_text(color = "blue"),
    axis.text.y.left = ggplot2::element_text(color = "blue"),
    axis.ticks.y.left = ggplot2::element_line(color = "blue"),
    axis.title.y.right = ggplot2::element_text(color = "red"),
    axis.text.y.right = ggplot2::element_text(color = "red"),
    axis.ticks.y.right = ggplot2::element_line(color = "red")
  ) +
  ggplot2::labs(x = "Precision", y = "Time of computation (ms)")
```
